cardinalR: Generating interesting high-dimensional data structures

### 2025-09-08 paper-cardinalR.Rmd
Stash

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The `geozoo` package (@barret2016) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is `snedata` (@james2025), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, `mlbench` (@friedrich2024) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, `splatter` (@luke2017) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression.

There is an  opportunity to improve the generation of high-dimensional data structures by integrating geometric principles with advanced noise control and customizable clustering. The `geozoo` package provides a strong foundation but could be enhanced to support high-dimensional extensions with controlled noise and user-defined parameters for clustering. Similarly, while `snedata` focuses on abstract datasets for dimensionality reduction, adding features for generating high-dimensional data from geometric layouts would enhance its usability. The `mlbench` package could also benefit from allowing users to create datasets with specific geometric structures and noise profiles. Additionally, although `splatter` specializes in biological data simulation, it could be expanded to offer a broader framework for generating diverse geometric structures across dimensions, enabling detailed control over noise and clustering. Addressing these areas could lead to more robust high-dimensional data generation tools.

To address this gap, this paper introduces the `cardinalR` R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. `cardinalR` offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By providing these features, `cardinalR` provides researchers with a method to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

The paper is organized as follows. In next section, introduces the implementation of `cardinalR` package on CRAN and GitHub, including demonstration of the package's key functions. We illustrate how a clustering data structure affect the dimension reductions in **Application** section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for use of our data collection.

>>>>

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The `geozoo` package (@barret2016) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is `snedata` (@james2025), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, `mlbench` (@friedrich2024) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, `splatter` (@luke2017) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression. While these packages are valuable, their scope is often limited to specific applications or low-dimensional structures.

To address this gap, this paper introduces the `cardinalR` R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. `cardinalR` offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By using clearly defined geometric shapes and controllable features such as number of dimensions, sample size; `cardinalR` allows researchers to generate transparent and interpretable synthetic datasets for evaluating algorithm performance in high-dimensional settings.

By providing these features, `cardinalR` provides researchers with a method to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

The paper is organized as follows. In the next section, we introduce the implementation of the `cardinalR` package on GitHub, including a demonstration of the package's key functions. We illustrate how a clustering data structure affects the dimension reductions in the Application section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for the use of our data collection.

### 2025-09-08 paper-cardinalR.tex
Stash

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The \texttt{geozoo} package (\citet{barret2016}) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is \texttt{snedata} (\citet{james2025}), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, \texttt{mlbench} (\citet{friedrich2024}) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, \texttt{splatter} (\citet{luke2017}) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression.

@@

The paper is organized as follows. In next section, introduces the implementation of \texttt{cardinalR} package on CRAN and GitHub, including demonstration of the package's key functions. We illustrate how a clustering data structure affect the dimension reductions in \textbf{Application} section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for use of our data collection.

>>>>

Several R packages address this need from different perspectives. For example, \texttt{geozoo} (\citet{barret2016}) provides a wide range of geometric objects, primarily in low dimensions; \texttt{snedata} (\citet{james2025}) focuses on datasets for dimensionality reduction; \texttt{mlbench} (\citet{friedrich2024}) supplies benchmark datasets for classification and regression; and \texttt{splatter} (\citet{luke2017}) specializes in simulating single-cell omics data. While each serves its own domain well, none are designed for flexible, high-dimensional data generation with customizable geometry, noise, and clustering.

@@

The paper is organized as follows. In the next section, we introduce the implementation of the \texttt{cardinalR} package on GitHub, including a demonstration of the package's key functions. We illustrate how a clustering data structure affects the dimension reductions in the Application section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for the use of our data collection.
